Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-24 paper.tex
Cites for phasing and imputation

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
Larger genome variation datasets are also being generated for other organisms
and a variety of purposes including agriculture~\citep{wang2023rice}, conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use 
bespoke binary formats~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate} 
that support the required access patterns.
While PLINK and BGEN support efficient access to genotype data, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits 
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

Processing of Biobank scale datasets can be split into a 
few broad categories. The most basic analysis 
is quality control (QC). Variant QC is an 
involved and multi-faceted 
task~\cite{marees2018tutorial,panoutsopoulou2018quality}, 
often requiring interactive, exploratory analysis
and incurring substantial computation over multiple QC fields.
Genotype calls are often refined via statistical methods,
% methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
for example by phasing~[CITES] and imputation[CITES], 
creating additional dataset copies.
A common task to perform after QC is 
a genome wide association study (GWAS)~\cite{uffelmann2021genome}. 
The majority of tools for 
performing GWAS and related analyses require
data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
and so data must be ``hard-called'' according to some QC criteria
and exported to additional copies.
Finally, variation datasets are often queried in exploratory
analyses, to find regions or samples of interest for a particular 
study~\cite[e.g.][]{chen2024novo}.

@@

Finally, to illustrate performance on a common filtering task, we 
created a copy of the VCF chunk which contains only variants 
that pass some common filtering criteria using
\texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
following standard practices [citations?].
This took a total of 689 min CPU time, with the same elapsed time.
In comparison, computing and storing a variant mask (i.e., a boolean value 
for each variant denoting whether it should be considered or not
for analysis) based on the same criteria using Zarr 
required a total of 1.96 min CPU time with an elapsed time 
of  11 min (63X faster).
%689 / 11 = 62.6 = 63X
There is an important distinction here between creating a copy
of the data (an implicit part of VCF based workflows) and creating an
additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
illustrates, call-level masks are cheap (the standard 
genotype missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
storage) and variant or sample level masks require negligible storage.
If downstream software can 
use configurable masks (at variant, sample and call level) 
rather than expecting full copies of the data, major storage savings
and improvements in processing efficiency can be made.
The transition from the manifold inefficiencies of 
present-day ``copy-oriented'' computing,
to the ``mask-oriented'' analysis of large immutable, single-source
datasets is a potentially transformational change enabled by Zarr.

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
Large genome variation datasets are also being generated for other organisms
and a variety of purposes including 
agriculture~\citep{ros2020accuracy,wang2023rice}, 
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats 
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits 
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

Processing of Biobank scale datasets can be split into a 
few broad categories. The most basic analysis 
is quality control (QC). Variant QC is an 
involved and multi-faceted 
task~\cite{marees2018tutorial,panoutsopoulou2018quality,chen2024genomic}, 
often requiring interactive, exploratory analysis
and incurring substantial computation over multiple QC fields.
Genotype calls are often refined via statistical methods,
for example by phasing~\citep{
browning2021fast,browning2023statistical,hofmeister2023accurate,williams2024phasing},
and imputation~\citep{browning2018one,rubinacci2020genotype,barton2021whole,
rubinacci2023imputation}
creating additional dataset copies.
A common task to perform after QC is 
a genome wide association study (GWAS)~\cite{uffelmann2021genome}. 
The majority of tools for 
performing GWAS and related analyses require
data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
and so data must be ``hard-called'' according to some QC criteria
and exported to additional copies.
Finally, variation datasets are often queried in exploratory
analyses, to find regions or samples of interest for a particular 
study~\cite[e.g.][]{chen2024novo}.

@@

Finally, to illustrate performance on a common filtering task, we 
created a copy of the VCF chunk which contains only variants 
that pass some common filtering criteria using
\texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
following standard practices~\citep[e.g.][]{chen2024genomic}.
This took a total of 689 min CPU time, with the same elapsed time.
In comparison, computing and storing a variant mask (i.e., a boolean value 
for each variant denoting whether it should be considered or not
for analysis) based on the same criteria using Zarr 
required a total of 1.96 min CPU time with an elapsed time 
of  11 min (63X faster).
%689 / 11 = 62.6 = 63X
There is an important distinction here between creating a copy
of the data (an implicit part of VCF based workflows) and creating an
additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
illustrates, call-level masks are cheap (the standard 
genotype missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
storage) and variant or sample level masks require negligible storage.
If downstream software can 
use configurable masks (at variant, sample and call level) 
rather than expecting full copies of the data, major storage savings
and improvements in processing efficiency can be made.
The transition from the manifold inefficiencies of 
present-day ``copy-oriented'' computing,
to the ``mask-oriented'' analysis of large immutable, single-source
datasets is a potentially transformational change enabled by Zarr.
